---
title: "Neuroendocrine RNA-seq Metastasis Project"
author: "Tyler Hearnsberger"
date: "2025-12-13"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

# 1. Project Overview

This project analyzes publicly available RNA-seq data from GEO accession
GSE98894, which profiles gene expression in gastro-entero-pancreatic
neuroendocrine tumors (GEP-NETs).

The main goal is to determine whether metastatic progression in GEP-NETs is
associated with a shared transcriptional program across pancreatic
neuroendocrine tumors (PanNETs) and small-intestinal neuroendocrine tumors
(SI-NETs), while avoiding confounding from tissue-of-origin differences.

```{r setup, message=FALSE, warning=FALSE}
# ============================================================
# 0. LIBRARIES
# ============================================================

library(DESeq2)
library(tidyverse)
library(apeglm)
library(pheatmap)
library(clusterProfiler)
library(org.Hs.eg.db)
```

# 2. Load Raw Counts

```{r load-counts, message=FALSE}
# ============================================================
# 1. LOAD RAW COUNTS
# ============================================================

counts <- read_tsv(
  "data/GSE98894_raw_counts_GRCh38.p13_NCBI.tsv",
  show_col_types = FALSE
)

counts <- counts %>%
  column_to_rownames("GeneID")
```

# 3. Parse GEO Metadata

```{r parse-metadata}
# ============================================================
# 2. LOAD RAW GEO METADATA
# ============================================================

meta_raw <- readLines("data/GSE98894_series_matrix.txt")

# ============================================================
# 3. EXTRACT GSM SAMPLE IDS
# ============================================================

gsm_line <- meta_raw[grep("^!Sample_geo_accession", meta_raw)]
gsm_ids <- strsplit(gsm_line, "\t")[[1]][-1]
gsm_ids <- gsub('"', "", gsm_ids)

# ============================================================
# 4. EXTRACT SAMPLE CHARACTERISTICS
# ============================================================

char_lines <- meta_raw[grep("!Sample_characteristics", meta_raw)]

site_line <- char_lines[
  grep("pancreas|intestin|ileum|rect", char_lines, ignore.case = TRUE)
][1]

site_vals <- strsplit(site_line, "\t")[[1]][-1]
site_vals <- gsub('"', "", site_vals)
site_vals <- tolower(site_vals)

status_line <- char_lines[
  grep("primary|metast", char_lines, ignore.case = TRUE)
][1]

status_vals <- strsplit(status_line, "\t")[[1]][-1]
status_vals <- gsub('"', "", status_vals)
status_vals <- tolower(status_vals)
```

# 4. Build Clean Metadata Table

```{r clean-metadata}
# ============================================================
# 5. BUILD CLEAN METADATA TABLE
# ============================================================

meta_clean <- tibble(
  sample = gsm_ids,
  site_raw = site_vals,
  status_raw = status_vals
) %>%
  mutate(
    site = case_when(
      str_detect(site_raw, "pancreas") ~ "PanNET",
      str_detect(site_raw, "intestin|ileum") ~ "SI-NET",
      str_detect(site_raw, "rect") ~ "RE-NET",
      TRUE ~ NA_character_
    ),
    status = case_when(
      str_detect(status_raw, "primary") ~ "Primary",
      str_detect(status_raw, "metast") ~ "Metastasis",
      TRUE ~ NA_character_
    )
  ) %>%
  select(sample, site, status)

# Focus project on PanNET and SI-NET samples.
meta_clean <- meta_clean %>%
  filter(site %in% c("PanNET", "SI-NET"))
```

# 5. Align Metadata and Counts

```{r align-data}
# ============================================================
# 6. ALIGN METADATA AND COUNTS
# ============================================================

common_samples <- meta_clean$sample

counts <- counts[, common_samples]
meta_clean <- meta_clean %>%
  arrange(match(sample, colnames(counts)))

stopifnot(all(meta_clean$sample == colnames(counts)))

# Sanity checks
dim(counts)
dim(meta_clean)
table(meta_clean$site, meta_clean$status)
```

# 6. Create DESeq2 Object

```{r create-dds}
# ============================================================
# 7. CREATE DESEQ2 OBJECT
# ============================================================

dds <- DESeqDataSetFromMatrix(
  countData = counts,
  colData   = meta_clean,
  design    = ~ site + status
)

dds <- dds[rowSums(counts(dds)) > 10, ]
dds
```

# 7. PCA Quality Control

PCA was used as an exploratory quality-control step. The goal was to determine
whether samples cluster according to known biological variables, especially
tissue of origin and metastatic status.

```{r vst}
# ============================================================
# 8. VARIANCE-STABILIZING TRANSFORMATION
# ============================================================

vsd <- vst(dds, blind = FALSE)
vsd_mat <- assay(vsd)
```

```{r global-pca, fig.width=7, fig.height=5}
# ============================================================
# 9. PCA - ALL SAMPLES
# ============================================================

pca_data <- plotPCA(vsd, intgroup = c("site", "status"), returnData = TRUE)
percentVar <- round(100 * attr(pca_data, "percentVar"))

p_global <- ggplot(pca_data, aes(PC1, PC2, color = site, shape = status)) +
  geom_point(size = 3, alpha = 0.8) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_classic() +
  ggtitle("PCA of GEP-NET Samples (VST)")

p_global

ggsave("figures/global_pca.png", p_global, width = 7, height = 5, dpi = 300)
```

## PCA Interpretation

The global PCA is expected to show tissue of origin as the dominant source of
variation, with PanNET and SI-NET samples separating primarily along PC1. This
supports the need for tissue-stratified analysis rather than pooling all tumors
together.

# 8. Tissue-Specific PCA

```{r pannet-pca, fig.width=7, fig.height=5}
# ============================================================
# 10. PCA - PanNET ONLY
# ============================================================

pan_idx <- meta_clean$site == "PanNET"
pca_pan <- prcomp(t(vsd_mat[, pan_idx]), scale. = FALSE)

pan_df <- data.frame(
  pca_pan$x,
  status = meta_clean$status[pan_idx]
)

p_pan <- ggplot(pan_df, aes(PC1, PC2, color = status)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic() +
  ggtitle("PCA - PanNET Only")

p_pan

ggsave("figures/pannet_pca.png", p_pan, width = 7, height = 5, dpi = 300)
```

```{r sinet-pca, fig.width=7, fig.height=5}
# ============================================================
# 11. PCA - SI-NET ONLY
# ============================================================

si_idx <- meta_clean$site == "SI-NET"
pca_si <- prcomp(t(vsd_mat[, si_idx]), scale. = FALSE)

si_df <- data.frame(
  pca_si$x,
  status = meta_clean$status[si_idx]
)

p_si <- ggplot(si_df, aes(PC1, PC2, color = status)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic() +
  ggtitle("PCA - SI-NET Only")

p_si

ggsave("figures/sinet_pca.png", p_si, width = 7, height = 5, dpi = 300)
```

# 9. PanNET Differential Expression

```{r pannet-de}
# ============================================================
# 12. DESEQ2 - PanNET ONLY
# ============================================================

dds_pan <- dds[, dds$site == "PanNET"]
dds_pan$site <- droplevels(dds_pan$site)
dds_pan$status <- relevel(dds_pan$status, ref = "Primary")

design(dds_pan) <- ~ status

dds_pan <- DESeq(dds_pan)

res_pan <- results(dds_pan, contrast = c("status", "Metastasis", "Primary"))
res_pan <- lfcShrink(
  dds_pan,
  coef = "status_Metastasis_vs_Primary",
  res = res_pan,
  type = "apeglm"
)

res_pan_df <- as.data.frame(res_pan) %>%
  rownames_to_column("GeneID") %>%
  arrange(padj)

write_csv(res_pan_df, "results/PanNET_DE_Metastasis_vs_Primary.csv")

summary(res_pan)
sum(res_pan_df$padj < 0.05, na.rm = TRUE)
```

```{r pannet-volcano, fig.width=7, fig.height=5}
# ============================================================
# 13. VOLCANO PLOT - PanNET
# ============================================================

res_pan_df <- res_pan_df %>%
  mutate(
    sig = if_else(
      !is.na(padj) & padj < 0.05 & abs(log2FoldChange) >= 1,
      "Significant",
      "Not significant"
    )
  )

p_volcano_pan <- ggplot(res_pan_df, aes(x = log2FoldChange, y = -log10(padj), color = sig)) +
  geom_point(alpha = 0.6, size = 1.5) +
  theme_classic() +
  ggtitle("PanNET: Metastasis vs Primary") +
  xlab("log2 Fold Change (Metastasis / Primary)") +
  ylab("-log10 adjusted p-value")

p_volcano_pan

ggsave("figures/pannet_volcano.png", p_volcano_pan, width = 7, height = 5, dpi = 300)
```

```{r pannet-top10}
top10_pan <- res_pan_df %>%
  filter(!is.na(padj)) %>%
  slice(1:10)

top10_pan
```

# 10. PanNET Heat Map

```{r pannet-heatmap, fig.width=8, fig.height=8}
# ============================================================
# 14. HEAT MAP - TOP PAN-NET METASTASIS GENES
# ============================================================

top_genes_pan <- res_pan_df %>%
  filter(!is.na(padj), padj < 0.05, abs(log2FoldChange) >= 1) %>%
  slice(1:40) %>%
  pull(GeneID)

vsd_pan <- vst(dds_pan, blind = FALSE)
mat_pan <- assay(vsd_pan)[top_genes_pan, ]

mat_pan_z <- t(scale(t(mat_pan)))

annotation_col <- data.frame(
  Status = colData(dds_pan)$status
)

rownames(annotation_col) <- colnames(mat_pan_z)

pheatmap(
  mat_pan_z,
  annotation_col = annotation_col,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_colnames = FALSE,
  fontsize_row = 8,
  scale = "none",
  main = "PanNET: Metastasis vs Primary\nTop Differentially Expressed Genes",
  filename = "figures/pannet_heatmap.png"
)
```

# 11. SI-NET Differential Expression

```{r sinet-de}
# ============================================================
# 15. DESEQ2 - SI-NET ONLY
# ============================================================

dds_sinet <- dds[, dds$site == "SI-NET"]
dds_sinet$site <- droplevels(dds_sinet$site)
dds_sinet$status <- relevel(dds_sinet$status, ref = "Primary")

design(dds_sinet) <- ~ status

dds_sinet <- DESeq(dds_sinet)

res_sinet <- results(dds_sinet)

res_sinet_df <- as.data.frame(res_sinet) %>%
  rownames_to_column("GeneID") %>%
  arrange(padj)

write_csv(res_sinet_df, "results/SINET_DE_Metastasis_vs_Primary.csv")

summary(res_sinet)
sum(res_sinet_df$padj < 0.05, na.rm = TRUE)
```

# 12. Shared Metastasis Signature

```{r shared-signature}
# ============================================================
# 16. SHARED SIGNIFICANT GENES
# ============================================================

padj_cutoff <- 0.05

res_pan_clean <- as.data.frame(res_pan)
res_sinet_clean <- as.data.frame(res_sinet)

res_pan_clean <- res_pan_clean[!is.na(res_pan_clean$padj), ]
res_sinet_clean <- res_sinet_clean[!is.na(res_sinet_clean$padj), ]

pan_sig <- res_pan_clean[res_pan_clean$padj < padj_cutoff, ]
sinet_sig <- res_sinet_clean[res_sinet_clean$padj < padj_cutoff, ]

pan_genes <- rownames(pan_sig)
sinet_genes <- rownames(sinet_sig)

shared_genes <- intersect(pan_genes, sinet_genes)

pan_fc <- pan_sig[shared_genes, "log2FoldChange"]
sinet_fc <- sinet_sig[shared_genes, "log2FoldChange"]

same_direction <- sum(sign(pan_fc) == sign(sinet_fc))
opposite_direction <- sum(sign(pan_fc) != sign(sinet_fc))

table_shared_summary <- data.frame(
  Metric = c(
    "PanNET significant genes",
    "SI-NET significant genes",
    "Shared significant genes",
    "Percent of PanNET shared",
    "Percent of SI-NET shared",
    "Shared genes same direction",
    "Shared genes opposite direction"
  ),
  Value = c(
    length(pan_genes),
    length(sinet_genes),
    length(shared_genes),
    round(100 * length(shared_genes) / length(pan_genes), 2),
    round(100 * length(shared_genes) / length(sinet_genes), 2),
    same_direction,
    opposite_direction
  )
)

table_shared_summary

write_csv(table_shared_summary, "results/shared_signature_summary.csv")
```

```{r shared-fc-plot, fig.width=7, fig.height=5}
plot_df <- data.frame(
  PanNET = pan_fc,
  SINET = sinet_fc
)

p_shared_fc <- ggplot(plot_df, aes(x = PanNET, y = SINET)) +
  geom_point(alpha = 0.6, size = 1.5) +
  theme_classic() +
  xlab("log2FC PanNET (Metastasis vs Primary)") +
  ylab("log2FC SI-NET (Metastasis vs Primary)") +
  ggtitle("Shared Metastasis Genes: PanNET vs SI-NET")

p_shared_fc

ggsave("figures/shared_gene_log2fc_scatter.png", p_shared_fc, width = 7, height = 5, dpi = 300)
```

# 13. GO Enrichment of Shared Upregulated Metastasis Genes

```{r go-enrichment}
# ============================================================
# 17. GO ENRICHMENT
# ============================================================

shared_up_met <- shared_genes[
  pan_sig[shared_genes, "log2FoldChange"] > 0 &
    sinet_sig[shared_genes, "log2FoldChange"] > 0
]

length(shared_up_met)

entrez_genes <- as.character(shared_up_met)

ego_shared <- enrichGO(
  gene          = entrez_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)

ego_df <- as.data.frame(ego_shared)

write_csv(ego_df, "results/shared_upregulated_GO_enrichment.csv")

ego_df[, c("Description", "GeneRatio", "p.adjust", "Count")][1:10, ]
```

```{r go-dotplot, fig.width=8, fig.height=5}
p_go <- dotplot(
  ego_shared,
  showCategory = 8,
  font.size = 12
) +
  ggtitle("Shared Metastasis Program Enriched for Biological Processes") +
  theme(
    axis.text.y = element_text(size = 11),
    plot.title = element_text(hjust = 0.5, size = 15, face = "bold")
  )

p_go

ggsave("figures/shared_GO_dotplot.png", p_go, width = 8, height = 5, dpi = 300)
```

# 14. Shared Metastasis Signature Score

```{r signature-score, fig.width=7, fig.height=5}
# ============================================================
# 18. SIGNATURE SCORE
# ============================================================

sig_mat <- vsd_mat[rownames(vsd_mat) %in% shared_up_met, ]
sig_mat_z <- t(scale(t(sig_mat)))

signature_score <- colMeans(sig_mat_z, na.rm = TRUE)

sig_df <- data.frame(
  sample = colnames(vsd_mat),
  signature_score = signature_score,
  status = colData(dds)$status,
  site = colData(dds)$site
)

p_sig_all <- ggplot(sig_df, aes(x = status, y = signature_score, fill = status)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.8, alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "Shared Metastasis Signature Score",
    y = "Signature Score",
    x = ""
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none"
  )

p_sig_all

ggsave("figures/shared_signature_score_all.png", p_sig_all, width = 7, height = 5, dpi = 300)

t.test(signature_score ~ sig_df$status)
```

```{r signature-score-by-tissue, fig.width=8, fig.height=5}
p_sig_site <- ggplot(sig_df, aes(x = status, y = signature_score, fill = status)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.8, alpha = 0.6) +
  facet_wrap(~ site) +
  theme_minimal() +
  labs(
    title = "Shared Metastasis Signature Score by Tissue",
    y = "Signature Score",
    x = ""
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none"
  )

p_sig_site

ggsave("figures/shared_signature_score_by_tissue.png", p_sig_site, width = 8, height = 5, dpi = 300)
```

# 15. Ranked Shared Metastasis Gene List

```{r ranked-gene-list}
# ============================================================
# 19. BUILD RANKED SHARED METASTASIS GENE LIST
# ============================================================

padj_cutoff <- 0.05
lfc_cutoff <- 1

res_pan_df2 <- as.data.frame(res_pan) %>%
  rownames_to_column("GeneID")

res_sinet_df2 <- as.data.frame(res_sinet) %>%
  rownames_to_column("GeneID")

pan_sig2 <- res_pan_df2 %>%
  filter(!is.na(padj), padj < padj_cutoff, abs(log2FoldChange) >= lfc_cutoff)

sinet_sig2 <- res_sinet_df2 %>%
  filter(!is.na(padj), padj < padj_cutoff, abs(log2FoldChange) >= lfc_cutoff)

shared_tbl <- inner_join(
  pan_sig2 %>% select(GeneID, log2FoldChange, padj),
  sinet_sig2 %>% select(GeneID, log2FoldChange, padj),
  by = "GeneID",
  suffix = c("_PanNET", "_SINET")
)

shared_up_tbl <- shared_tbl %>%
  filter(log2FoldChange_PanNET > 0, log2FoldChange_SINET > 0) %>%
  mutate(
    mean_log2FC = (log2FoldChange_PanNET + log2FoldChange_SINET) / 2,
    mean_padj_score = ((-log10(padj_PanNET)) + (-log10(padj_SINET))) / 2
  ) %>%
  arrange(desc(mean_log2FC), desc(mean_padj_score))

gene_map <- bitr(
  shared_up_tbl$GeneID,
  fromType = "ENTREZID",
  toType = c("SYMBOL", "GENENAME"),
  OrgDb = org.Hs.eg.db
)

shared_up_annotated <- shared_up_tbl %>%
  left_join(gene_map, by = c("GeneID" = "ENTREZID")) %>%
  relocate(SYMBOL, GENENAME, .after = GeneID)

write_csv(shared_up_tbl, "results/shared_metastasis_gene_list_ranked.csv")
write_csv(shared_up_annotated, "results/shared_metastasis_gene_list_annotated.csv")

head(shared_up_annotated, 20)
```

# 16. Final Interpretation

This project supports a tissue-stratified view of metastasis in GEP-NETs.
Tissue of origin is the dominant source of transcriptomic variation, but both
PanNETs and SI-NETs show metastasis-associated gene expression changes. The
overlap of significant genes with consistent directionality suggests that a
shared metastatic transcriptional program exists across these tumor types.

At the same time, the PCA results indicate that metastasis is not simply one
uniform global shift across all samples. Instead, metastasis appears to be a
context-dependent process layered onto tissue-specific tumor biology.
